
---
title: "Week 11: Effects of Spatial Dependence"
output: github_document
---

```{r global_options, eval=T, echo=F,results='hide', error=F, warning=FALSE}

knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/',  warning=FALSE, message=FALSE, results='hide')
rm(list=ls())
library(spdep)
library(rgdal)
library(maptools)
library(RColorBrewer)
```


# Modeling areal data

## Spatial statistical approach

### Simultaneous autoregressive model (SAR)

The SAR specification uses a regression on the values from the other areas to
account for the spatial dependence. This means that the error terms $e$ are
modelled so that they depend on each other in the following way:

$$e_i=\sum_{i=1}^{m} b_{ij}e_i+\epsilon_i$$ 

where $\epsilon_i \sim N(0, \sigma_i^2)$

or $$e_1, e_2, \ldots, e_n \sim N(0, (I-B)^{-1}D((I-B)^{-1})^T) $$

If denote $e_i=Y-X^T\beta$, then the SAR model could be expressed as:

$Y-X^T\beta = B(Y-X^T\beta)+\epsilon$ or $(I-B)(Y-X^T\beta)=\epsilon$ 


where $B$ is a matrix that contains the dependence parameters $b_{ij}$ and $I$
is the identity matrix of the required dimension. It is important to point out
that in order for this SAR model to be well defined, the matrix $I − B$ must be
non-singular. 

A common practice to specify $B$ is $B=\rho W$, where $W$ is the spatial weight
matrix and $\rho$ is the `spatial autoregression parameter`.

Frequently employed in the spatial econometric literature with two variants,
spatial lag model and spatial error model. 


### Conditional autoregressive model (CAR)

The CAR specification relies on the conditional distribution of the spatial
error terms. In this case, the distribution of $e_i$ conditioning on $e_{−i}$ is
given.

$$e_i|e_{-i} \sim N(\sum_{j}b_{ij}e_j, \tau_i^2)$$

where $j$ is the neighbors of $i$.

Using Brook’s Lemma we can obtain:

$$e_1, e_2, \ldots, e_n \sim N(0, (I-B)^{-1}D)$$


Compared to SAR model, the CAR model is easier to be fitted into a Bayesian 
hierarchical models. 


```{r area2, eval=T, echo=T}

# display the maps 

NY8 <- readOGR("Data", "NY8_utm18")
TCE <- readOGR("Data", "TCE")
cities <- readOGR("Data", "NY8cities")

par(mfrow=c(1,2))
plot(NY8, border="grey60", axes=TRUE)
text(coordinates(cities), labels=as.character(cities$names), font=2, cex=0.9)
text(bbox(NY8)[1,1], bbox(NY8)[2,2], labels="a)", cex=0.8)

plot(NY8, border="grey60", axes=TRUE)
points(TCE, pch=1, cex=0.7)
points(TCE, pch=3, cex=0.7)
text(coordinates(TCE), labels=as.character(TCE$name), cex=0.7,
 font=1, pos=c(4,1,4,1,4,4,4,2,3,4,2), offset=0.3)
text(bbox(NY8)[1,1], bbox(NY8)[2,2], labels="b)", cex=0.8)

# ordinary linear regression

nylm <- lm(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME, data=NY8)
summary(nylm)
NY8$lmresid <- residuals(nylm)
NY_nb <- read.gal("Data/NY_nb.gal", region.id=row.names(NY8))
NYlistw<-nb2listw(NY_nb, style = "B")

# Moran's I
lm.morantest(nylm, NYlistw)


# Spatial conditional and simultaneous autoregression 

nysar<-spautolm(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME, data=NY8, listw=NYlistw)

summary(nysar)


nylam1 <- c(nysar$lambda)
nylam2 <- c(LR1.spautolm(nysar)$p.value)
# Notice above that there is still strong spatial autocorrelation, and the proximty to TCE seems not quite significant

# Display the trend and residual component
NY8$sar_trend <- nysar$fit$signal_trend
NY8$sar_stochastic <- nysar$fit$signal_stochastic
rds <- colorRampPalette(brewer.pal(8, "RdBu"))
tr_at <- seq(-1, 1.3, length.out=21)
tr_rds <- rds(sum(tr_at >= 0)*2)[-(1:(sum(tr_at >= 0)-sum(tr_at < 0)))]
tr_pl <- spplot(NY8, c("sar_trend"), at=tr_at, col="transparent", col.regions=tr_rds, main=list(label="Trend", cex=0.8))
st_at <- seq(-0.16, 0.39, length.out=21)
st_rds <- rds(sum(st_at >= 0)*2)[-(1:(sum(st_at >= 0)-sum(st_at < 0)))]
st_pl <- spplot(NY8, c("sar_stochastic"), at=st_at, col="transparent", col.regions=st_rds, main=list(label="Stochastic", cex=0.8))
plot(tr_pl, split=c(1,1,2,1), more=TRUE)
plot(st_pl, split=c(2,1,2,1), more=FALSE)

# The proximity to a TCE seems not to be significant, after we include
# the population as weights, it becomes significant.

nylmw <- lm(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME, data=NY8, weights=POP8)
summary(nylmw)
NY8$lmwresid <- residuals(nylmw)

# Display
gry <- c(rev(brewer.pal(6, "Reds")[1:4]), colorRampPalette(brewer.pal(5, "Blues"))(9))
TCEpts <- list("sp.points", TCE, pch=16, col="grey5")
spplot(NY8, c("lmresid", "lmwresid"), sp.layout=list(TCEpts), col.regions=gry, col="transparent", lwd=0.5, at=seq(-2,4.5,0.5))

# Now check the moran's again

lm.morantest(nylmw, NYlistw)

# Include weights in spautolm

nysarw<-spautolm(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME , data=NY8, listw=NYlistw, weights=POP8)
summary(nysarw)

NY8$sarw_trend <- nysarw$fit$signal_trend
NY8$sarw_stochastic <- nysarw$fit$signal_stochastic
tr_pl <- spplot(NY8, c("sarw_trend"), at=tr_at, col="transparent", col.regions=tr_rds, main=list(label="Trend", cex=0.8))
st_pl <- spplot(NY8, c("sarw_stochastic"), at=st_at, col="transparent", col.regions=st_rds, main=list(label="Stochastic", cex=0.8))
plot(tr_pl, split=c(1,1,2,1), more=TRUE)
plot(st_pl, split=c(2,1,2,1), more=FALSE)

# Conditional Autoregressive Models

nycar<-spautolm(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME , data=NY8, family="CAR",
   listw=NYlistw)
summary(nycar)



nycarw<-spautolm(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME, data=NY8, family="CAR",
   listw=NYlistw, weights=POP8)
summary(nycarw)



# SMA model 
#nysmaw<-spautolm(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME, data=NY8, family="SMA",
#   listw=NYlistw, weights=POP8)
#summary(nysmaw)

# Spatial economitrics approach

#NYlistwW <- nb2listw(NY_nb, style = "W")
#res <- lm.LMtests(nylm, listw=NYlistwW, test="all")
#tres <- t(sapply(res, function(x) c(x$statistic, x$parameter, x$p.value)))
#colnames(tres) <- c("Statistic", "df", "p-value")
#printCoefmat(tres)
#
#
#nylag <- lagsarlm(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME, data=NY8, listw=NYlistwW)
#summary(nylag)
#bptest.sarlm(nylag)
#
#
#
#McRes <- sarml(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME, wmat=listw2mat(NYlistwW), eigvar=eigenw(NYlistwW), print=FALSE, data=NY8)
#c(McRes$beta, rho=McRes$rho, sig2=McRes$sig2)
#
#
#nymix <- lagsarlm(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME, data=NY8, listw=NYlistwW, type="mixed")
#nymix
#anova(nymix, nylag)
#
#
#W <- as(as_dgRMatrix_listw(NYlistwW), "CsparseMatrix")
#trMat <- trW(W, type="mult")
#head(trMat)
#
#
#
#
#nyerr <- errorsarlm(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME, data=NY8, listw=NYlistwW)
#summary(nyerr)
#LR.sarlm(nyerr, nymix)
#
#
#nyerr1 <- errorsarlm(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME, data=NY8, listw=NYlistwW, etype="emixed")
#coef(nyerr1)
#

```
